Augmentation of a Historical Harpsichord Keyboard Replica for Haptic-Enabled Interaction in Museum Exhibitions

### 2025-02-04 src/article.tex

The \anon{Tagliavini} Collection in \anon{Bologna \cite{Tagliavini2007}}, renowned for its historical keyboard instruments, exemplifies this dilemma. Housed at Museo San Colombano - Genus Bononiae, with more than seventy early keyboard instruments, primarily early plucked stringed keyboards of italian origin. The collection stands out as a valuable resource for musicologists, organologists and musicians alike. Preserving the instruments' authenticity was the cornerstone of \anon{Ferdinando Tagliavini}’s vision. This guiding principle led him to collect instruments which could be restored to playing condition after minimal intervention.

@@

As part of the initiative to improve the accessibility of museums, the Museo San Colombano - Tagliavini Collection, part of the cultural itinerary of Fondazione Carisbo, has designed a series of projects to improve cognitive and sensory accessibility at different levels, in a project funded by NextGenerationEU. The aim of this project is to exploit the potential of the collection of musical instruments to create an immersive experience and interaction with the heritage on display. 
reinterpretation of the historical musical objects in their tactile-auditory dimension aims to open up the exhibition to blind and partially sighted visitors. The augmented keyboard fulfils many of the project's aims and is part of the initiative to allow visitors of different abilities to engage in a hands-on creative process within the museum.

@@

The keyboard presented here inherits some aspects from the Tromba Moderna project. The project aims to offer a tool to enhance the fruition of the \anon{Tagliavini} collection while retaining a form of continuity with historical instrument-building traditions. However, the electronics hidden inside the interface are not intended to augment or disrupt the tactile feedback; rather, they serve as the silent and invisible link between the mechanical and the digital realms. 
The optical sensing technique for the keyboard is adapted from a similar project on the piano by McPherson \cite{McPherson2013}. 
Thus, whether the interface proposed here is a \emph{new} musical interface is subject to discussion, particularly regarding the keyboard's tactile response designed to adhere to longstanding harpsichord building traditions. This work complements and follows up on previous reflections within the NIME community, emphasising the \emph{O} in NIME -- the methodologies and reasoning for exploring `old' instrument designs at NIME \cite{Masu_NIME_2023}. 
\textcolor{red}{ALT: This work has incorporated and developed further the ideas of reusing and complimenting the `old' in digital musical instrument design as put forward in `The \emph{O} in NIME' \cite{Masu_NIME_2023}}.
The intended use of the keyboard through meaningful interaction with a museum exhibit is where the novelty of this work lies, rather than solely in its technological development. Relying on previous NIMEs, this project extends the longevity of the results beyond the scope envisaged in the original works, enhancing their sustainability, as discussed by Masu \emph{et al.} \cite{Masu_NIME_2023}.

@@

% In an effort to make the exhibition more accessible and authentic, the \anon{San Colombano Museum} commissioned the construction of a mechanical keyboard to trigger a MIDI-controlled sample library, designed for use by museum visitors. 
The exhibition is set to open approximately a month after this work's submission, and only preliminary feedback has been collected from a pool of 20 people -- consisting of staff, surveillance personnel and visitors who were given training on the system -- and expert feedback from \anon{Catalina Vicens} and \anon{Roberto Livi}.
The collection of visitor feedback was carried out as short informal interviews and primarily as a means of identifying any technical problems.
The keyboard shows promise in enhancing the museum experience as comments about a ``sense of disconnect'' found in the Benton Fletcher Collection \cite{McAlpine2014} were absent in this initial feedback.

@@

A more formal surveying of museum visitors is to be carried out to determine to what degree the exhibition achieved its core aims.
The keyboard is hosted in the \emph{Oratory} above the museum's main hall, an exceptional testimony to the Bolognese art, decorated by the finest students of the Carracci and presenting a series of frescoes covering most of the walls and ceiling. The room hosts unique examples of the Italian Renaissance building tradition, including the 1547 harpsichord and the 1540 spinet by \anon{Alessandro Trasuntino}. Visitors often visit the room with the same caution and respect typical of worship spaces. Among the visitors interviewed, no comments were made that the exposed headphones and touchscreen affected such an experience negatively, though a more definitive answer will come as more feedback is collected after launch. Pictures of the keyboard hosted in the \emph{Oratory} are shown in Figure \ref{fig:oratory}.

>>>>

The \anon{Tagliavini} Collection in \anon{Bologna \cite{Tagliavini2007}}, renowned for its historical keyboard instruments, exemplifies this dilemma. Housed at \anon{Museo San Colombano - Genus Bononiae}, with more than seventy early keyboard instruments, primarily early plucked stringed keyboards of Italian origin. The collection stands out as a valuable resource for musicologists, organologists and musicians alike. Preserving the instruments' authenticity was the cornerstone of \anon{Ferdinando Tagliavini}’s vision. This guiding principle led him to collect instruments which could be restored to playing condition after minimal intervention.

@@

\textcolor{red}{
As part of the initiative to improve the accessibility of museums, the Museo San Colombano - Tagliavini Collection, part of the cultural itinerary of Fondazione Carisbo, has designed a series of projects to improve cognitive and sensory accessibility at different levels, in a project funded by NextGenerationEU. The aim of this project is to exploit the potential of the collection of musical instruments to create an immersive experience and interaction with the heritage on display. 
reinterpretation of the historical musical objects in their tactile-auditory dimension aims to open up the exhibition to blind and partially sighted visitors. The augmented keyboard fulfils many of the project's aims and is part of the initiative to allow visitors of different abilities to engage in a hands-on creative process within the museum.
}

@@

The keyboard presented here inherits some aspects from the Tromba Moderna project. The project aims to offer a tool to enhance the fruition of the \anon{Tagliavini} collection while retaining a form of continuity with historical instrument-building traditions. However, the electronics hidden inside the interface are not intended to augment or disrupt the tactile feedback; rather, they serve as the silent and invisible link between the mechanical and the digital realms. 
The optical sensing technique for the keyboard is adapted from a similar project on the piano by McPherson \cite{McPherson2013}. 
Thus, whether the interface proposed here is a \emph{new} musical interface is subject to discussion, particularly regarding the keyboard's tactile response designed to adhere to longstanding harpsichord building traditions. This work has incorporated and developed further the ideas of reusing and complimenting the `old' in digital musical instrument design as put forward in `The \emph{O} in NIME' \cite{Masu_NIME_2023}.
The intended use of the keyboard through meaningful interaction with a museum exhibit is where the novelty of this work lies, rather than solely in its technological development. This project extends the longevity of the results beyond the scope envisaged in the original works, enhancing their sustainability, as discussed by Masu \emph{et al.} \cite{Masu_NIME_2023}.

@@

The exhibition is set to open approximately a month after this work's submission, and only preliminary feedback has been collected from a pool of 20 people -- consisting of staff, surveillance personnel and visitors who were given training on the system -- and expert feedback from \anon{Catalina Vicens} and \anon{Roberto Livi}.
The collection of visitor feedback was carried out as short informal interviews and primarily as a means of identifying any technical problems.
The keyboard shows promise in enhancing the museum experience as comments about a ``sense of disconnect'' found in the Benton Fletcher Collection \cite{McAlpine2014} were absent in this initial feedback.

@@

A more formal surveying of museum visitors would be required to determine to what degree the exhibition achieved its core aims.
The keyboard is hosted in the \emph{Oratory} above the museum's main hall, an exceptional testimony to the Bolognese art, decorated by the finest students of the Carracci and presenting a series of frescoes covering most of the walls and ceiling. The room hosts unique examples of the Italian Renaissance building tradition, including the 1547 harpsichord and the 1540 spinet by \anon{Alessandro Trasuntino}. Visitors often visit the room with the same caution and respect typical of worship spaces. Among the visitors interviewed, no comments were made that the exposed headphones and touchscreen affected such an experience negatively, though a more definitive answer will come as more feedback is collected after launch. Pictures of the keyboard hosted in the \emph{Oratory} are shown in Figure \ref{fig:oratory}.
